### 2015-09-16 revise/rebuttal150916.txt

#Reviewer 1

The authors have declined to incorporate the changes that I suggested. As a consequence, I can only standby my original comments and leave the final decision to the Editor. This is somewhat disappointing as I feel the additional data would substantially elevate the authors work. I would remark that if no additions are to be made then the authors should qualify the claims in the discussion somewhat. Whilst adding additional features into the model is straightforward, the authors present no evidence to indicate that the resultant analysis would yield anything sensible.

—————
As the reviewer says, performing small analysis incorporating some epigenetic data without careful validation is “straightforward” and not so hard. We have newly included simulation study, the result of which indicate that adding epigenetic features will help improving the accuracy of mutation signature detection as well as gaining novel biological insights.

There are several issues to be addressed for incorporating epigenetic data to mutation signature on real data. First, the choice of normalization methods of epigenetic data will strongly influence the result. Also, there are known relationships between epigenetic features and sequencing depth and errors (replication timing, CG contents) leading the variation in the sensitivity and false positive ratios. So we need to carefully control them.

In spite of the above obstacles, many studies have partly revealed that somatic mutations (mostly structural variations so far) have some relationships with epigenetic features and people will start to try clarifying more comprehensive relationships between mutation signatures and epigenetic data in the not-so-distant future. We believe our proposed model, which gives a natural way for incorporating many additional features, will be of great help in that analysis.
—————

caption for S7 Figure:
The accuracy of the proposed approach for the simulated data when changing the number of features starting from the case where just substation patterns and immediate 5’ and 3’ bases(M = (6, 4, 4), to 5 additional features (M = (6, 4, 4, 2, 2, 2, 2, 2) for each number of samples, dispersion parameters.  The accuracy of the estimated mutation signature improves as the number of additional features increases, indicating that incorporating additional features such as epigenetic data will be potentially beneficial.

### 2015-09-18 paper/manuscript/plos_template.tex
rebutal

%There were several problems in these classical studies \cite{pmid12379884, pmid15748635}.
The potential for classical studies to yield insights into somatic mutation processes
were limited in several ways.
Due to limited sequencing throughputs, most classical studies 
focused on a few cancer genes, such as TP53,
where high mutation frequencies could be expected. They then
contrasted mutation pattern profiles among different cancer types,
aggregating mutations across multiple individuals within the same cancer type to
yield sufficient mutations for analysis.
However, since many of the mutations in cancer genes are driver mutations causing cell proliferation,
the resultant mutation profiles are a biased representation of the underlying mutation process.
Furthermore, the paucity of mutation data made it effectively impossible to assess variation in mutation patterns among individuals.

@@

To make the most of these opportunities requires the development
of efficient and effective statistical methods for analyzing
mutation signatures in vast amounts of somatic mutation data.
Current statistical approaches \cite{pmid23318258,pmid23628380}, 
%based on nonnegative matrix factorization (NMF, \cite{pmid10548103}) and Poisson models
are an excellent starting point, 
and have helped generate the new insights noted above.
However, we argue here that these existing methods 
have two important limitations, caused by the fact that they
use an unconstrained model for each ``mutation signature."
First, although using an unconstrained model might appear to be a good thing in terms of flexibility, 
in practice it can actually reduce flexibility, because
the price of using an unconstrained model is that one must limit 
the domain of mutation signatures considered. 
For example, most recent analyses of mutation signatures
consider only the immediate flanking 5' and 3' bases of each substitution to be part of
the signature, even though it is known that more distal bases -- and particularly the
next flanking base on each side -- can contain important contextual information
\cite{pmid9683596}. These recent analyses take this approach because, in the unconstrained model,
incorporating the more distal bases into the signature very substantially increases the number of parameters, making estimated mutation signatures unstable.
Secondly, and just as important, the unconstrained model means
that each signature is a probability distribution in a high-dimensional
parameter space, which can make signatures difficult to interpret.

In this paper, we present a novel probabilistic approach 
to mutation signature modelling that addresses these limitations.
In brief, we first simplify the modelling of mutation signatures by decomposing them into separate ``mutation features".
For example, the substitution type is one feature; flanking bases are each another feature. We then exploit this 
decomposition by using a probabilistic model for signatures that assumes independence across features. This approach substantially reduces the number of parameters associated with
each signature, greatly facilitating the incorporation of additional relevant sequence context. 
For example, our approach can
incorporate both the two bases
3' and 5' of the substitution, and transcription strand biases using only 18 parameters
per signature, compared with 3071 parameters per signature with current approaches. We demonstrate the benefits of this simplification in data analyses. These benefits include: more stable estimation of signatures from smaller samples, 
refinement of the detail and resolution of many mutation signatures, and, possibly, identification of novel signatures.

Assuming independence among features in a signature may initially seem unnatural. However, its use here is analogous  ``position weight matrix models" which have been highly
successful 
for modelling transcription factor binding motifs.
Indeed, an important second contribution of
our paper is to provide intuitive visual representations for mutation signatures,
analogous to the ``sequence logos" used for visualizing binding motifs.
Finally, we also highlight the close connection
between mutation signature models and the ``mixed-membership models'',
also known as``admixture models" \cite{pmid10835412} or ``latent Dirichlet allocation''
models \cite{Blei:2003} that are widely used in population genetics 
and document clustering applications.
These connections should be helpful for future elaboration of 
computational and statistical methods for cancer mutation signature detection.

@@

The first contribution of this paper is to suggest
a more parsimonious approach to modelling 
mutation signatures, with the benefit of producing both more stable estimates and more easily-interpretable signatures.
In brief, we substantially reduce the number of parameters
per signature by breaking each mutation pattern into 
``features", and assuming independence
across mutation features. For example, consider
the case where a mutation pattern is defined by the substitution and its two flanking bases. We break this into three features (substitution,
3' base, 5' base), and characterize each mutation
signature by a probability distribution for each feature
(which, by our independence assumption, are multiplied
together to define a distribution on mutation patterns). Since the number of possible values for each feature is 6, 4, and 4 respectively this requires $5+3+3=11$ parameters instead of $96-1=95$ parameters. Furthermore, extending this model to account for $\pm n$ neighboring bases requires only $5 + 6n$ parameters instead of $6 \times 4^{2n} - 1$. For example,
 considering $\pm 2$ positions requires 17 parameters
 instead of 1535. Finally, incorporating
transcription strand as an additional feature adds just one parameter, instead of doubling the number of parameters.

Since the aim of a mutation signature is, in some ways, to capture dependencies among features, the independence assumption may seem counter-intuitive. However, the
idea is exactly analogous to the use of a ``position weight matrix" (PWM) to represent motifs in sequence data. In this
analogy, a motif is analogous to a mutation signature, and each location in the motif is analogous to a ``feature". Just as we use a probability vector for each feature, a PWM defines a probability vector for each location in a motif,
and these probabilities at each location can be multiplied together to yield a probability
distribution on sequences.
Even though a PWM cannot capture complex 
higher-order dependencies, some of which likely do exist in practice, it has been a highly successful tool for motif
analysis - likely because it can capture the most important characteristics of transcription factor binding sites 
(that some locations will show strong preference for a particular base, whereas others will not), and also
because it can be represented in an easily-interpretable way via sequencing logos \cite{pmid2172928}.
For similar reasons -- in addition to the empirical
demonstrations we present later -- we believe our mutation signature representation will prove useful for mutation signature analysis.

Fig. 1 illustrates the way that our new representation of signatures can simply capture
a previously-identified signature \cite{pmid25362482, pmid23318258} and provides
an easily-interpretable visualization of the signature 
that is analogous to  sequencing logos \cite{pmid2172928}.
We particularly note how the key elements of this mutation signature are more immediately visually apparent than with visualizations of the full vector of probabilities used by existing approaches.

@@

\caption{{\bf Examples of visualizations and parameter values for the mutation signatures of the unconstrained (full) model and our independent model, 
where substitution patterns, two 5' and 3' bases and transcription strand direction are considered as mutation features.}
(A) The barplots are divided by 6 substitution patterns and transcription strand direction.
In each division, 256 bars show joint probabilities of up to two base 5' and 3' bases
(ApApNpApA, ApApNpApC, ApApNpApG, ApApNpApT, $\cdots$, TpTpNpTpT).
(B, C) An example mutation signature representation and parameter values from our independent model, where mutation features (substitution patterns, two 5' and 3' bases and strand direction) 
are assumed to be independent ($L=6$, $\bm{M} = (6, 4, 4, 4, 4, 2)$).
In the bottom five rectangles, the width of each box represents the frequencies of bases (A, C, G and T) at the substitution and flanking site. 
To highlight the most informative flanking sites, the heights of flanking site boxes are scaled by ($1 + 0.5 \times \log \sum_{n = A,C,G,T} f_n^2$, where $f_n$ is the parameter for each base), which can be interpreted as ($1 - 0.5 \times \text{R\'enyi entropy}$) \cite{rrnyi1961measures}. This is analogous to the information content scaling used in sequencing logos.
In the top rectangle, the height of each box represents the conditional frequencies of mutated bases for each original base (C and T).
In the upper right, the height of the $+$ box represents the frequencies of mutations in the coding strand (or the plus strand, the sense strand and the untranscribed strand)
whose nucleotide sequences directly corresponds to mRNA, 
whereas the height of $-$ box represents those in the template strand (or the minus strand, the antisense strand, the transcribed strand and the noncoding strand)
whose sequences are copied during the synthesis of mRNA.
}
\label{mutSig_example}

@@

The data consist of 14717 somatic substitutions 
collected from a study of 26 urothelial carcinomas of the upper urinary tract (UCUT) \cite{pmid23926200}.
The original study identified a novel mutation signature in these data: 
T $>$ A substitutions at CpTpG sites with a strong transcription strand specificity caused by aristolochic acids (AA).

We consider a mutation pattern to consist of the substitution pattern, the $\pm 2$ flanking bases, and the transcription strand direction. Thus each signature is characterized by 18 parameters in our independent model, and by 3071 parameters in the full model. After analyzing the data with various number of mutation signatures $K$,
we selected $K=3$ signatures for these analyses (see S2 Text).

The inferred APOBEC signature under the independent model
shows a clear depletion of G base at the $-2$ position,
which is consistent to the previous study \cite{pmid23318258} and results in the next subsection (Figs. 3A and 3B).
In contrast, for the full model, this tendency is rather mild (Figs. 3C, 3D and S1).
The inferred AA mutation signature has no clear characteristics at the $-2$ position.
These results suggest that our independent model has the potential
to identify signatures in more detail and with less data than existing approaches based on the full model.

@@

To provide a more comprehensive practical illustration of
our method, we applied it to somatic mutation data from 30 cancer types \cite{pmid23945592}.
We applied the method to each cancer type separately to assess similarity of estimated signatures across cancer types.
For each cancer type we selected the number of signatures $K$ by fitting the model with increasing $K$ and examining the log-likelihood, bootstrap errors, and correlation of membership parameters.
The selected values of $K$ are given in S2 Table.
Also, we simply removed somatic mutations located in an intergenic region to include transcription strand biases as mutation features.
Finally, we merged similar mutation signatures across different cancer types 
(when their Frobenius Distance were $<$ $0.6$). 
% XXX WHAT KIND OF CLUSTERING?
 % (see Supplementary Material for the values of the log-likelihoods and bootstrap errors for each cancer type).

Figs. \ref{nature2013_sig_summary} and \ref{nature2013_sig_member} summarise the results.
In total, we identified 27 mutation signatures.
Many of these signatures show reassuring similarities with
signatures identified in previous studies. However
signatures from our independence model, because of its ability to effectively
and parsimoniously deal with both $\pm2$ flanking
base context and strand bias, are often more refined, highlighting additional details or features not previously evident. By comparing the composition of nucleotides and cancer types exhibiting the signatures with results of previous studies, we were able to associate many of the detected signatures with known mutational processes.
In addition, as we reviewed these signatures
and compared them with previous work, we noticed connections
that, while not directly related to our new model, appear
novel and noteworthy. The remainder of this section
provides a comprehensive discussion of these findings.

@@

Signatures 1 and 8 (C $>$ A at TpCpT and C $>$ T at TpCpG, respectively) observed in colorectal and uterine cancers
appear likely to be associated with deregulated activity of the error-prone polymerase Pol $\epsilon$. 
In previous analyses of these data \cite{pmid23945592}, 
the signature for Pol $\epsilon$ dysfunction was represented by a single signature (their ``signature 10'').
In contrast our new approach uses two signatures.
Since these signatures are highly correlated,
and appear connected by
a single biological mechansim, we certainly do not argue
that inferring them as a single signature is ``wrong". However, splitting them into two signatures does help highlight certain features. Specifically,
signature 1 shows a transcription strand bias whereas signature 8 does not, and this is true for both colorectal and uterus cancers (Figs. S2C S2D).
This strand bias may be connected with the enrichment of 
C $>$A at TpCpT mutations in leading strands of replication forks
observed by \cite{pmid25228659}. Although replication strand bias is different from transcription strand bias, these two biases may be connected through the fact that replication origins prefer transcription start sites \cite{pmid23187890}.

These signatures also illustrate the ability of our model
to help highlight sequence context effects beyond the immediate flanking bases. Specifically, both signatures 1 and 8  show an elevated frequency of the T base at position $-2$,
and signature 1 also shows slightly elevated frequency of the T base at position $+2$
 (Figs. 6B, 6C ,S6C and S6D).
A previous study of Pol $\epsilon$  \cite{pmid25228659} 
found that a nonsense mutation R23X of TP53 is enriched in cancers with Pol $\epsilon$ defects.
In fact, the pattern of this mutation is C $>$ T at TpTpCpGpA, closely matching signature 8.
This illustrates that the inclusion of $\pm 2$ bases 
into signatures may be helpful for identifying underlying mechanisms.

\begin{figure}[h]
% \subfigure[APOBEC signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=14.5cm,height=6cm]{APOBEC_two5prime.eps}
%    \label{APOBEC_two5prime}}
% \subfigure[POLE1 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE1_two5prime.eps}
%   \label{POLE1_two5prime}}
% \quad
% \subfigure[POLE2 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE2_two5prime.eps}
%   \label{POLE2_two5prime}}
% \subfigure[UV signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{UV_two5prime.eps}
%   \label{UV_two5prime}}
% \quad
% \subfigure[The signature 11 intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{LMST_two5prime.eps}
%   \label{LMST_two5prime}}
\caption{{\bf The estimated frequencies of bases at two 5' to the mutated site for each cancer type.}
The bar heights show the estimated frequency for bases A, C, G and T at two 5' to the mutated site.
The error bars show bootstrapped standard errors.
(A, B, C, D, E) The intensities of signature 13 (APOBEC signature), signature 1 (the first POL $\epsilon$ signature),
signature 8 (the second POL $\epsilon$ signature), signature 10 (ultraviolet signature) and signature 11, respectively, at two 5' to the mutated site.
intensities at two 5' to the mutated site.
}
\label{two5prime}
\end{figure}

@@

%XX this is important that it is drive by a single sample, but maybe just because yuichi looked more carefully and not new method XX
Signature 2 (C $>$ A at [CT]pCpT) is observed solely in low grade gliomas, and
appears related to, but slightly different from, the signature previously detected in the same cancer types (``signature 14'', \cite{pmid23318258}). 
Indeed, the corresponding signature in the previous study shows very complex patterns (C $>$ A at NpCpT or C $>$ T at GpTpN). Further investigation
revealed that this signature is driven by 
a single sample with an extremely high mutation rate (see Figs. S3A and S3B), and signature 2 disappeared when we removed this sample (see Fig. S4).
It may be that the complex low-grade-glioma specific signature detected in the previous study is driven by the same single sample.
We suggest that these signatures should be treated with caution until validated in additional samples. 
% XXX Need to discuss this wording and what it means XXX

%XX Also important, making a connection, but not the new method XX
Signature 4 (C $>$ A at CpCpG) observed in kidney clear cell carcinomas, lung adenocarcinomas and melanomas 
seems to correspond to the ``signature R2''  detected in the same cancer types (plus lung squamous carcinomas) in 
\cite{pmid23318258} (see their Supplementary Figures).
Again our analysis highlights additional contextual information, with a strikingly elevated frequency of base C at the $-2$ position (Figs. S5A, S5B and S5C).
However, for each cancer type, only a few samples support this signature (see Figs. S5D, S5E and S5F),
and the corresponding signature 
could not be validated in the previous study: most somatic mutations corresponding to that signature could not be validated by re-sequencing or visual inspection of BAM files using genomic viewers. Again, further investigation
yielded a potential explanation for this finding:
this signature largely matches that of a putative artifact caused by oxidation of DNA during acoustic shearing \cite{pmid23303777}, 
and we conclude that this signature, and the corresponding signature in previous work, are likely artefactual.
Although not of direct biological interest,
identifying artefactual signatures could be helpful in removing false positive mutations.

%XX highlights robustness of approach XX
Signature 13 (T $>$ [AGT] at TpCpN sites) was observed in 12 cancer types, and is surely related to the activity of the APOBEC family. The 12 inferred signatures are highly consistent among cancer types except for B-cell lymphoma (see Fig. S2A), highlighting the robustness
of our approach.  Almost all of them show enrichment of A and T and depletion of G base at the $-2$ position (Fig. 6A and S2A), 
consistent with the UCUT data above and previous analyses \cite{pmid23318258}.
The estimated transcribed strand specificities varied among cancer types, suggesting that there is not consistent strand-specificity in APOBEC signatures (and the observed variation may be due to estimation errors).
%Yuichi is that true?
Signatures 15 and 16 may also be related to APOBEC, although the estimated signatures are sufficiently different from 13 that they
were not merged into a single 
cluster by our specified clustering criteria.

>>>>

%There were several problems in these classical studies \cite{pmid12379884, pmid15748635}.
The potential for classical studies to yield insights into somatic mutation processes
were limited in several ways.
Due to limited sequencing throughput, most classical studies 
focused on a few cancer genes, such as TP53,
where high mutation frequencies could be expected. They then
contrasted mutation pattern profiles among different cancer types,
aggregating mutations across multiple individuals within the same cancer type to
yield sufficient mutations for analysis.
However, since many of the mutations in cancer genes are driver mutations causing cell proliferation,
the resultant mutation profiles are a biased representation of the underlying mutation process.
Furthermore, the paucity of mutation data made it effectively impossible to assess variation in mutation patterns among individuals.

@@

To make the most of these opportunities requires the development
of efficient and effective statistical methods for analyzing
mutation signatures in vast amounts of somatic mutation data.
Current statistical approaches \cite{pmid23318258,pmid23628380}
%based on nonnegative matrix factorization (NMF, \cite{pmid10548103}) and Poisson models
are an excellent starting point, 
and have helped generate the new insights noted above.
However, we argue here that these existing methods 
have two important limitations, caused by the fact that they
use an unconstrained model for each ``mutation signature."
First, although using an unconstrained model might appear to be a good thing in terms of flexibility, 
in practice it can actually reduce flexibility, because
the price of using an unconstrained model is that one must limit 
the domain of mutation signatures considered. 
For example, most recent analyses of mutation signatures
consider only the immediate flanking 5' and 3' bases of each substitution to be part of
the signature, even though it is known that more distal bases -- and particularly the
next flanking base on each side -- can contain important contextual information
\cite{pmid9683596}. These recent analyses take this approach because, in the unconstrained model,
incorporating the more distal bases into the signature very substantially increases the number of parameters, making estimated mutation signatures unstable.
Secondly, and just as important, the unconstrained model means
that each signature is a probability distribution in a high-dimensional
parameter space, which can make signatures difficult to interpret.

In this paper, we present a novel probabilistic approach 
to mutation signature modelling that addresses these limitations.
In brief, we first simplify the modelling of mutation signatures by decomposing them into separate ``mutation features".
For example, the substitution type is one feature; flanking bases are each another feature. We then exploit this 
decomposition by using a probabilistic model for signatures that assumes independence across features. This approach substantially reduces the number of parameters associated with
each signature, greatly facilitating the incorporation of additional relevant sequence context. 
For example, our approach can incorporate the two bases 3' and 5' of the substitution and transcription strand biases using only 18 parameters
per signature, compared with 3071 parameters per signature with current approaches. We demonstrate the benefits of this simplification in data analyses. These benefits include: more stable estimation of signatures from smaller samples, 
refinement of the detail and resolution of many mutation signatures, and possibly identification of novel signatures.

Assuming independence among features in a signature may initially seem unnatural. However, its use here is analogous to ``position weight matrix models" which have been highly
successful 
for modelling transcription factor binding motifs.
Indeed, an important second contribution of
our paper is to provide intuitive visual representations for mutation signatures,
analogous to the ``sequence logos" used for visualizing binding motifs.
Finally, we also highlight the close connection
between mutation signature models and the ``mixed-membership models'',
also known as``admixture models" \cite{pmid10835412} or ``latent Dirichlet allocation''
models \cite{Blei:2003} that are widely used in population genetics 
and document clustering applications.
These connections should be helpful for future elaboration of 
computational and statistical methods for cancer mutation signature detection.

@@

The first contribution of this paper is to suggest
a more parsimonious approach to modelling 
mutation signatures, with the benefit of producing both more stable estimates and more easily interpretable signatures.
In brief, we substantially reduce the number of parameters
per signature by breaking each mutation pattern into 
``features", and assuming independence
across mutation features. For example, consider
the case where a mutation pattern is defined by the substitution and its two flanking bases. We break this into three features (substitution,
3' base, 5' base), and characterize each mutation
signature by a probability distribution for each feature
(which, by our independence assumption, are multiplied
together to define a distribution on mutation patterns). Since the number of possible values for each feature is 6, 4, and 4 respectively this requires $5+3+3=11$ parameters instead of $96-1=95$ parameters. Furthermore, extending this model to account for $\pm n$ neighboring bases requires only $5 + 6n$ parameters instead of $6 \times 4^{2n} - 1$. For example,
 considering $\pm 2$ positions requires 17 parameters
 instead of 1535. Finally, incorporating
transcription strand as an additional feature adds just one parameter, instead of doubling the number of parameters.

Since the aim of a mutation signature is, in some ways, to capture dependencies among features, the independence assumption may seem counter-intuitive. However, the
idea is exactly analogous to the use of a ``position weight matrix" (PWM) to represent motifs in sequence data. In this
analogy, a motif is analogous to a mutation signature, and each location in the motif is analogous to a ``feature". Just as we use a probability vector for each feature, a PWM defines a probability vector for each location in a motif,
and these probabilities at each location can be multiplied together to yield a probability
distribution on sequences.
Even though a PWM cannot capture complex 
higher-order dependencies, some of which likely do exist in practice, it has been a highly successful tool for motif
analysis - likely because it can capture the most important characteristics of transcription factor binding sites 
(that some locations will show strong preference for a particular base, whereas others will not), and also
because it can be represented in an easily interpretable way via sequencing logos \cite{pmid2172928}.
For similar reasons -- in addition to the empirical
demonstrations we present later -- we believe our mutation signature representation will prove useful for mutation signature analysis.

Fig. 1 illustrates the way that our new representation of signatures can simply capture
a previously identified signature \cite{pmid25362482, pmid23318258} and provides
an easily interpretable visualization of the signature 
that is analogous to  sequencing logos \cite{pmid2172928}.
We particularly note how the key elements of this mutation signature are more immediately visually apparent than with visualizations of the full vector of probabilities used by existing approaches.

@@

\caption{{\bf Examples of visualizations and parameter values for the mutation signatures of the unconstrained (full) model and our independent model, 
where substitution patterns, two 5' and 3' bases and transcription strand direction are considered as mutation features.}
(A) The barplots are divided by 6 substitution patterns and transcription strand direction.
In each division, 256 bars show joint probabilities of up to two base 5' and 3' bases
(ApApNpApA, ApApNpApC, ApApNpApG, ApApNpApT, $\cdots$, TpTpNpTpT).
(B, C) An example mutation signature representation and parameter values from our independent model, where mutation features (substitution patterns, two 5' and 3' bases and strand direction) 
are assumed to be independent ($L=6$, $\bm{M} = (6, 4, 4, 4, 4, 2)$).
In the bottom five rectangles, the width of each box represents the frequencies of bases (A, C, G and T) at the substitution and flanking site. 
To highlight the most informative flanking sites, the heights of flanking site boxes are scaled by 
$1 + 0.5 \times \log \sum_{n = A,C,G,T} f_n^2$, where $f_n$ is the parameter for each base, 
which can be interpreted as $1 - 0.5 \times \text{R\'enyi entropy}$ \cite{rrnyi1961measures}. This is analogous to the information content scaling used in sequencing logos.
In the top rectangle, the height of each box represents the conditional frequencies of mutated bases for each original base (C and T).
In the upper right, the height of the $+$ box represents the frequencies of mutations in the coding strand (or the plus strand, the sense strand and the untranscribed strand)
whose nucleotide sequences directly corresponds to mRNA, 
whereas the height of $-$ box represents those in the template strand (or the minus strand, the antisense strand, the transcribed strand and the noncoding strand)
whose sequences are copied during the synthesis of mRNA.
}
\label{mutSig_example}

@@

The data consist of 14,717 somatic substitutions 
collected from a study of 26 urothelial carcinomas of the upper urinary tract (UCUT) \cite{pmid23926200}.
The original study identified a novel mutation signature in these data: 
T $>$ A substitutions at CpTpG sites with a strong transcription strand specificity caused by aristolochic acids (AA).

We consider a mutation pattern to consist of the substitution pattern, the $\pm 2$ flanking bases, and the transcription strand direction. Thus each signature is characterized by 18 parameters in our independent model, and by 3071 parameters in the full model. After analyzing the data with various numbers of mutation signatures $K$,
we selected $K=3$ signatures for these analyses (see S2 Text).

The inferred APOBEC signature under the independent model
shows a clear depletion of G base at the $-2$ position,
which is consistent with the previous study \cite{pmid23318258} and results in the next subsection (Figs. 3A and 3B).
In contrast, for the full model, this tendency is rather mild (Figs. 3C, 3D and S1).
The inferred AA mutation signature has no clear characteristics at the $-2$ position.
These results suggest that our independent model has the potential
to identify signatures in more detail and with less data than existing approaches based on the full model.

@@

To provide a more comprehensive practical illustration of
our method, we applied it to somatic mutation data from 30 cancer types \cite{pmid23945592}.
We applied the method to each cancer type separately to assess similarity of estimated signatures across cancer types.
For each cancer type we selected the number of signatures $K$ by fitting the model with increasing $K$ and examining the log-likelihood, bootstrap errors, and correlation of membership parameters.
The selected values of $K$ are given in S2 Table.
Also, we simply removed somatic mutations located in an intergenic region to include transcription strand biases as mutation features.
Finally, we merged similar mutation signatures across different cancer types 
(when their Frobenius distances were $<$ $0.6$, where the Frobenius distance between two matrices ($\bm{F}_1$, $\bm{F}_2$) is $\sqrt{ \Tr  \bigl( (\bm{F}_1 -  \bm{F}_2 )  (\bm{F}_1 -  \bm{F}_2 )^t \bigr) }$,
where $\Tr$ means the trace of square matrices.
% XXX WHAT KIND OF CLUSTERING?
 % (see Supplementary Material for the values of the log-likelihoods and bootstrap errors for each cancer type).

Figs. \ref{nature2013_sig_summary} and \ref{nature2013_sig_member} summarise the results.
In total, we identified 27 mutation signatures.
Many of these signatures show reassuring similarities with
signatures identified in previous studies. However
signatures from our independence model, because of its ability to effectively
and parsimoniously deal with both $\pm 2$ flanking
base context and strand bias, are often more refined, highlighting additional details or features not previously evident. By comparing the composition of nucleotides and cancer types exhibiting the signatures with results of previous studies, we were able to associate many of the detected signatures with known mutational processes.
In addition, as we reviewed these signatures
and compared them with previous work, we noticed connections
that, while not directly related to our new model, appear
novel and noteworthy. The remainder of this section
provides a comprehensive discussion of these findings.

@@

Signatures 1 and 8 (C $>$ A at TpCpT and C $>$ T at TpCpG, respectively) observed in colorectal and uterine cancers
appear likely to be associated with deregulated activity of the error-prone polymerase Pol $\epsilon$. 
In previous analyses of these data \cite{pmid23945592}, 
the signature for Pol $\epsilon$ dysfunction was represented by a single signature (their ``signature 10'', see Fig. S2).
In contrast our new approach uses two signatures.
Since these signatures are highly correlated,
and appear connected by
a single biological mechanism, we certainly do not argue
that inferring them as a single signature is ``wrong". However, splitting them into two signatures does help highlight certain features. Specifically,
signature 1 shows a transcription strand bias whereas signature 8 does not, and this is true for both colorectal and uterus cancers (Figs. S3C and S3D).
This strand bias may be connected with the enrichment of 
C $>$A at TpCpT mutations in leading strands of replication forks
observed by \cite{pmid25228659}. Although replication strand bias is different from transcription strand bias, these two biases may be connected through the fact that replication origins prefer transcription start sites \cite{pmid23187890}.

These signatures also illustrate the ability of our model
to help highlight sequence context effects beyond the immediate flanking bases. Specifically, both signatures 1 and 8  show an elevated frequency of the T base at position $-2$,
and signature 1 also shows slightly elevated frequency of the T base at position $+2$
 (Figs. 6B, 6C ,S3C and S3D).
A previous study of Pol $\epsilon$  \cite{pmid25228659} 
found that a nonsense mutation R23X of TP53 is enriched in cancers with Pol $\epsilon$ defects.
In fact, the pattern of this mutation is C $>$ T at TpTpCpGpA, closely matching signature 8.
This illustrates that the inclusion of $\pm 2$ bases 
into signatures may be helpful for identifying underlying mechanisms.

\begin{figure}[h]
% \subfigure[APOBEC signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=14.5cm,height=6cm]{APOBEC_two5prime.eps}
%    \label{APOBEC_two5prime}}
% \subfigure[POLE1 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE1_two5prime.eps}
%   \label{POLE1_two5prime}}
% \quad
% \subfigure[POLE2 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE2_two5prime.eps}
%   \label{POLE2_two5prime}}
% \subfigure[UV signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{UV_two5prime.eps}
%   \label{UV_two5prime}}
% \quad
% \subfigure[The signature 11 intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{LMST_two5prime.eps}
%   \label{LMST_two5prime}}
\caption{{\bf The estimated frequencies of bases at two 5' to the mutated site for each cancer type.}
The bar heights show the estimated frequency for bases A, C, G and T at the $-2$ position (two 5' to the mutated site).
The error bars show bootstrapped standard errors.
(A, B, C, D, E) The intensities of signature 13 (APOBEC signature), signature 1 (the first Pol $\epsilon$ signature),
signature 8 (the second Pol $\epsilon$ signature), signature 10 (ultraviolet signature) and signature 11, respectively, at the $-2$ position.
}
\label{two5prime}
\end{figure}

@@

%XX this is important that it is drive by a single sample, but maybe just because yuichi looked more carefully and not new method XX
Signature 2 (C $>$ A at [CT]pCpT) is observed solely in low grade gliomas, and
appears related to, but slightly different from, the signature previously detected in the same cancer types (``signature 14'', \cite{pmid23318258}). 
Indeed, the corresponding signature in the previous study shows very complex patterns (C $>$ A at NpCpT or C $>$ T at GpTpN). Further investigation
revealed that this signature is driven by 
a single sample with an extremely high mutation rate (see Figs. S4A and S4B), and signature 2 disappeared when we removed this sample (see Fig. S5).
It may be that the complex low-grade-glioma specific signature detected in the previous study is driven by the same single sample.
We suggest that these signatures should be treated with caution until validated in additional samples. 
% XXX Need to discuss this wording and what it means XXX

%XX Also important, making a connection, but not the new method XX
Signature 4 (C $>$ A at CpCpG) observed in kidney clear cell carcinomas, lung adenocarcinomas and melanomas 
seems to correspond to the ``signature R2''  detected in the same cancer types (plus lung squamous carcinomas) in 
\cite{pmid23318258} (see their Supplementary Figures).
Again our analysis highlights additional contextual information, with a strikingly elevated frequency of base C at the $-2$ position (Figs. S6A, S6B and S6C).
However, for each cancer type, only a few samples support this signature (see Figs. S6D, S6E and S6F),
and the corresponding signature 
could not be validated in the previous study: most somatic mutations corresponding to that signature could not be validated by re-sequencing or visual inspection of BAM files using genomic viewers. Again, further investigation
yielded a potential explanation for this finding:
this signature largely matches that of a putative artifact caused by oxidation of DNA during acoustic shearing \cite{pmid23303777}, 
and we conclude that this signature, and the corresponding signature in previous work, are likely artefactual.
Although not of direct biological interest,
identifying artefactual signatures could be helpful in removing false positive mutations.

%XX highlights robustness of approach XX
Signature 13 (T $>$ [AGT] at TpCpN sites) was observed in 12 cancer types, and is surely related to the activity of the APOBEC family. The 12 inferred signatures are highly consistent among cancer types except for B-cell lymphoma (see Fig. S3A), highlighting the robustness
of our approach.  Almost all of them show enrichment of A and T and depletion of G base at the $-2$ position (Fig. 6A and S3A), 
consistent with the UCUT data above and previous analyses \cite{pmid23318258}.
The estimated transcribed strand specificities varied among cancer types, suggesting that there is not consistent strand-specificity in APOBEC signatures (and the observed variation may be due to estimation errors).
%Yuichi is that true?
Signatures 15 and 16 may also be related to APOBEC, although the estimated signatures are sufficiently different from 13 that they
were not merged into a single 
cluster by our specified clustering criteria.

Signatures 10, 11, 12, 19 and 21 provide further examples
of our method refining previously observed signatures, highlighting strand biases and/or sequence context effects,
particularly 2 bases upstream of the substitution.
Signature 10 (C $>$T at [CT]pCpC) was observed in head and neck cancers and melanomas, and probably relates to ultraviolet light.
Consistent strand specificities among the two cancer types (Fig. S3E) matches previous results \cite{pmid23318258}, 
but our analysis additionally highlights elevated abundance of T at the $-2$ position (Figs. 6D and S3E). 
Signature 11 (C $>$ T at GpCp[CG]) appears in small-cell lung cancers and stomach cancers,
and seems to be the same as ``signature 15'' in the previous study, whose function remains unclear. Again our analysis highlights elevated abundance of G at the $-2$ position (Figs. 6E and S3F).
Signatures 12 (C $>$ T at [CG]pCp[CT]), 19 (T $>$ C at GpTpN) and 21 (T $>$ [CG] at CpTpT) observed in pilocytic astrocytomas, stomach cancers and oesophagus cancers, respectively, agree well with those detected in the same cancer types in the previous study \cite{pmid23318258}. However our analysis again refines these signatures,
highlighting a strand bias in all three, and sequence context effects at the -2 position in Signatures 12 and 21.
%although the causes of these signatures remain unknown.
%which should be validated by analysis of more samples from different cohorts and cancer types.
